---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# INDEED
## Overview


This package implements INDEED algorithm from Zuo et. al.'s Methods paper: INDEED: Integrated differential expression and differential network analysis of omic data for biomarker discovery (PMID: 27592383). 

This R package will generate a csv file containing information such as p-values, node degree and activity score for each biomolecule. A higher activity score indicates that the corresponding biomolecule has more neighbors connceted in the differential network and their p-values are more statistically significant. It will also generate a csv file for the differential network created by INDEED. 

## Installation

You can install INDEED from github with:

```{r gh-installation, eval=FALSE}
# The development version from GitHub:
# install.packages("devtools")
devtools::install_github("ressomlab/INDEED")
```

## Usage


```{r example, eval=FALSE}
library(INDEED)

# Example 1:
# Using partial correlation to obtain sparse differential network
pre_data <- pre_partial(data=Met_GU,class_label = Met_Group_GU,id=Met_name_GU)
partial_cor(data_list=pre_data,rho_group1='min',rho_group2="min",permutation = 1000,p_val=pvalue_M_GU)


# Example 2:
# Using Spearman correlation to obtain differential network
non_partial_cor(data=Met_GU,class_label = Met_Group_GU,id=Met_name_GU,method="spearman")
```


